---
title: "BayRC Workflow: Comparative Circadian Genomics"
author: "Thien Quy Pham"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{BayRC Workflow: Comparative Circadian Genomics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # all chunks set to eval=FALSE; run interactively with your data
)
```

> **Audience:** Biologists with RNA-seq experience; no Bayesian statistics
> background required. For the full mathematical derivations see
> Pham *et al.* (2025) and the package supplementary materials (`supplementary.tex`).

BayRC answers one question that standard differential expression tools cannot:
**how has the circadian transcriptome been remodeled between two biological
conditions?** Which genes switched from oscillating to flat? Which gained
rhythmicity? For genes that kept oscillating, do they still peak at the same
hour? BayRC propagates posterior uncertainty from the gene level through pathway
enrichment and genome-wide concordance scoring, so that confidence in a single
gene's rhythmicity directly feeds into all downstream analyses.

# Data Requirements and Setup

## Input format

BayRC needs two objects per condition:

- **Expression matrix** (`Y`): G genes × N samples, log-normalized (log-CPM,
  VST, or similar). Row names must be gene symbols.
- **Zeitgeber time vector** (`t`): length N, hours, matching column order of Y.

```{r}
dim(expr_PUT)    # [G, N], e.g., [15000, 48]
length(time_PUT) # must equal ncol(expr_PUT)
head(time_PUT)   # e.g., 0 2 4 6 8 10 ...
```

## Installation

```{r}
devtools::install("/path/to/BayRC")

# Bioconductor dependencies
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("fgsea")
```

# Step 1: MCMC Inference

For each gene, BayRC fits:
$$Y_g(t) = M_g + A_g \cos\!\left(\tfrac{2\pi}{24}(t - \phi_g)\right) + \varepsilon_g$$

where $M_g$ is the MESOR (midline), $A_g$ the amplitude, $\phi_g$ the acrophase
(peak time), and $\rho_g \in \{0,1\}$ a rhythmicity indicator. Reversible Jump
MCMC explores both the flat ($\rho_g = 0$) and rhythmic ($\rho_g = 1$) models
in a single chain, using slice sampling for efficient amplitude-phase updates.

## Running the sampler

```{r}
library(BayRC)

# --- Condition A: Putamen ---
data_list_PUT <- list(
  data  = as.data.frame(expr_PUT),
  time  = time_PUT,
  gname = rownames(expr_PUT)
)
init_PUT  <- CB_init_single(Data.list = data_list_PUT, P = 24)
mcmc_PUT  <- CB_MCMC_single_rj_slice(
  Data.list  = data_list_PUT,
  Init.value = init_PUT,
  P          = 24,
  iteration  = 3000,  # post-thinning stored iterations
  n.burn     = 1000,  # warm-up discarded
  thin       = 20,    # keep 1 of every 20 raw steps
  p_rhythmic = rep(0.2, nrow(data_list_PUT$data))  # prior P(rhythmic)
)

# --- Condition B: Substantia Nigra ---
data_list_SUN <- list(
  data  = as.data.frame(expr_SUN),
  time  = time_SUN,
  gname = rownames(expr_SUN)
)
init_SUN  <- CB_init_single(Data.list = data_list_SUN, P = 24)
mcmc_SUN  <- CB_MCMC_single_rj_slice(
  Data.list  = data_list_SUN,
  Init.value = init_SUN,
  P = 24, iteration = 3000, n.burn = 1000, p_rhythmic = rep(0.2, nrow(data_list_SUN$data))
)

# Save immediately — MCMC is the most expensive step
saveRDS(mcmc_PUT, "mcmc_PUT.rds")
saveRDS(mcmc_SUN, "mcmc_SUN.rds")
```

> **Output:** Each `mcmc_*` object contains G × K matrices: `$rho` (rhythmicity
> indicator), `$phi` (acrophase), `$A` (amplitude), `$M` (MESOR). K = stored
> iterations after thinning.

## Optional: convergence diagnostics

```{r}
mcmc_PUT <- CB_MCMC_single_rj_slice(
  Data.list = data_list_PUT, Init.value = init_PUT,
  P = 24, iteration = 3000, n.burn = 1000,
  p_rhythmic = rep(0.2, nrow(data_list_PUT$data)),
  diagnostics = TRUE   # prints ESS, acceptance rate; appends $diagnostics to output
)
# Console output:
# === MCMC Diagnostics ===
# Samples stored:          3000
# Mean acceptance rate:    0.135
# Mean ESS (rho):          842
```

> **Interpretation:** Acceptance rates of 10–20% indicate healthy RJMCMC mixing.
> Mean ESS > 200 is satisfactory for BFDR-controlled downstream analysis.

# Step 2: Annotate and Extract Summaries

## Mandatory annotation with `match_symbols()`

```{r}
# Attaches gene symbols and RHYindex (BF > 3 binary call) to the MCMC object.
# p_rhythmic must match the value used in CB_MCMC_single_rj_slice (default 0.2).
mcmc_PUT <- match_symbols(mcmc_PUT, BF = 3)
mcmc_SUN <- match_symbols(mcmc_SUN, BF = 3)
```

## Posterior rhythmicity probability

```{r}
pA <- rowMeans(mcmc_PUT$rho)   # P(rhythmic | data), condition A
pB <- rowMeans(mcmc_SUN$rho)   # condition B

hist(pA, breaks = 50, main = "P(rhythmic | data) — Putamen",
     xlab = "Posterior probability", col = "steelblue")
```

> **Interpretation:** `pA = 0.92` means 92% of MCMC samples supported the
> rhythmic model for that gene — a direct probability statement, not a p-value.

## Bayes Factor

```{r}
# p_rhythmic must match the MCMC prior (0.2) or BF values will be miscalibrated.
bf_PUT <- summarize_bay(mcmc_PUT$rho, BF = 3, p_rhythmic = 0.2)
bf_SUN <- summarize_bay(mcmc_SUN$rho, BF = 3, p_rhythmic = 0.2)

cat("PUT rhythmic genes (BF > 3):", sum(bf_PUT$Rhythmicity), "\n")
```

| BF | Evidence for rhythmicity |
|---|---|
| 1–3 | Weak |
| 3–10 | Moderate |
| 10–100 | Strong |
| > 100 | Decisive |

## Amplitude, phase, and credible intervals

```{r}
# CB_getAllEst extracts posterior medians and 95% circular HDIs for phi, A, M
est_PUT <- CB_getAllEst(mcmc_PUT, burn = 100)

# est_PUT[[1]] — phase: phi.Est, phi.Lower, phi.Upper
# est_PUT[[2]] — amplitude: A.Est, A.Lower, A.Upper
# est_PUT[[3]] — MESOR: M.Est, M.Lower, M.Upper
head(est_PUT[[1]][order(bf_PUT$BayesF, decreasing = TRUE), ])
```

> **Interpretation:** The phase credible interval (cHDI) is the shortest arc on
> the 24-hour circle containing 95% of posterior samples — correctly handling the
> ZT 0/24 wrap-around. A narrow interval means tightly constrained peak timing.

# Step 3: Transition Classification

```{r}
trans <- transition_classify(pA, pB, bfdr_alpha = 0.25)

# Prints thresholds and counts:
# τ_gain = 0.142 | n_gain = 312
# τ_loss = 0.118 | n_loss = 523
# τ_cons = 0.198 | n_cons = 677

table(trans$gain_loss_status)
# Non-rhythmic   Gain   Loss   Maintained
#        11934    312    523          677
```

Joint transition posteriors used:
- **P(gain)** = (1 − pA) × pB
- **P(loss)** = pA × (1 − pB)
- **P(maintained)** = pA × pB

> **Interpretation:** `gain_loss_status` is the central annotation vector. Every
> downstream step uses these labels. A Gain gene was flat in condition A and
> oscillates in condition B — the clock acquired a new transcriptional target. A
> Loss gene lost circadian regulation — a common signature of aging, disease, or
> metabolic stress. A Maintained gene oscillates in both: the substrate for phase
> analysis.

# Step 4: Phase Concordance

```{r}
phase <- phase_infer(
  phi_matrix1      = mcmc_PUT$phi,
  phi_matrix2      = mcmc_SUN$phi,
  gain_loss_status = trans$gain_loss_status,
  shift            = 2,          # δ = 2 h tolerance (paper primary threshold)
  P                = 24,
  bfdr_alpha       = 0.25,
  compute_hdi      = TRUE        # compute 95% circular HDI of Δφ
)

cat("Phase-conserved:", sum(phase$flag_cons,  na.rm = TRUE), "\n")
cat("Phase-shifted:  ", sum(phase$flag_shift, na.rm = TRUE), "\n")
```

> **Interpretation:**
>
> - `flag_cons = TRUE`: the gene oscillates in both conditions and peaks within
>   ±2 h of the same time. The circadian phase is biologically conserved.
> - `flag_shift = TRUE`: the gene oscillates in both but has shifted its peak
>   timing. Oscillation is maintained but the clock has been reset for this gene.
> - `deltaPhi.Est`: the posterior median of Δφ = φ_A − φ_B in hours. Positive
>   values mean condition A peaks earlier; negative values mean later.
> - A circular HDI [3.1, 7.4] that does not contain zero provides clear evidence
>   of a 3–7 hour advance in condition A relative to condition B.

# Step 5A: Two-Stage Pathway Enrichment

## Stage 1 — circadian-active pathways

```{r}
kegg <- readRDS("kegg_pathway_list_hsa.rds")

result_union <- pathSelect(
  mcmc.merge.list = list(A = mcmc_PUT, B = mcmc_SUN),
  pathway.list    = kegg,
  dataset.names   = c("PUT", "SUN"),
  ranking.method  = "union",   # score = −2 log[(1−pA)(1−pB)]
  score_type      = "pos",
  qvalue.cut      = 0.20,
  nperm           = 1000,
  seed            = 42
)

active_names        <- result_union$results$pathway[result_union$results$padj < 0.20]
active_pathway_list <- kegg[active_names]
cat(length(active_names), "pathways active in Stage 1\n")
```

## Stage 2 — transition enrichment within active pathways

```{r}
result_gain <- pathSelect(mcmc.merge.list = list(A = mcmc_PUT, B = mcmc_SUN),
  pathway.list = active_pathway_list, dataset.names = c("PUT", "SUN"),
  ranking.method = "gain", score_type = "pos", qvalue.cut = 0.20, nperm = 1000)

result_loss <- pathSelect(mcmc.merge.list = list(A = mcmc_PUT, B = mcmc_SUN),
  pathway.list = active_pathway_list, dataset.names = c("PUT", "SUN"),
  ranking.method = "loss", score_type = "pos", qvalue.cut = 0.20, nperm = 1000)

result_cons <- pathSelect(mcmc.merge.list = list(A = mcmc_PUT, B = mcmc_SUN),
  pathway.list = active_pathway_list, dataset.names = c("PUT", "SUN"),
  ranking.method = "conserved", score_type = "pos", qvalue.cut = 0.20, nperm = 1000)

nrow(result_gain$results[result_gain$results$padj < 0.20, ])
```

> **Gain-Loss Ratio (GLR):** `Gain_Loss_Ratio_Arithmetic` in the results table.
> GLR > 1 = pathway gained more rhythmicity than it lost in condition B.
> GLR < 1 = more loss than gain — pathway uncoupled from the clock.

# Step 5B: Genome-Wide Concordance (c-score)

```{r}
global <- multi_conservation(
  mcmc.merge.list     = list(A = mcmc_PUT, B = mcmc_SUN),
  dataset.names       = c("PUT", "SUN"),
  select.pathway.list = "global",
  n_perm              = 1000,
  n_boot              = 500,
  output.dir          = "results/concordance",
  use_cpp             = TRUE,
  save_output         = TRUE   # set FALSE to suppress Excel write in non-interactive use
)

cat("Adjusted c-score:", round(global$Adjusted["Global", "PUT&SUN"], 3), "\n")
cat("p-value:         ", round(global$PValue["Global", "PUT&SUN"],   4), "\n")
cat("Genome-wide GLR: ", round(global$Gain_loss_ratio["Global", "PUT&SUN"], 3), "\n")
```

> **Interpretation:**
>
> - **c-score > 0.3, p < 0.01:** strong transcriptome-wide circadian conservation.
> - **c-score ≈ 0.1–0.2, p < 0.05:** modest but significant conservation.
> - **c-score ≈ 0, p > 0.1:** two largely independent circadian programs.
>
> The genome-wide GLR summarizes the net direction: is the circadian system gaining
> or losing rhythmic output from condition A to condition B across the whole
> transcriptome?

# Biological Interpretation Checklist

After completing the analysis, work through these questions:

- [ ] **Rhythmicity landscape:** What fraction of genes is rhythmic in each
  condition (`pA > 0.5`, `pB > 0.5`)? A large difference suggests broad circadian
  dampening or activation.

- [ ] **Dominant transition:** Compare `n_gain` vs. `n_loss`. Loss-dominant =
  circadian uncoupling (common in aging, neurodegeneration). Gain-dominant =
  circadian recruitment of new gene targets.

- [ ] **Phase shifts:** Are shifts in `phase$deltaPhi.Est` systematic (all similar
  magnitude/direction, implying a global clock advance or delay) or gene-specific
  (implying altered transcription factor binding)?

- [ ] **Pathway biology:** Which pathways are enriched for gain vs. loss? Check
  `Top_Gain_Genes` and `Top_Loss_Genes` in pathway results to identify driver genes
  for follow-up validation.

- [ ] **Global concordance:** Is the c-score significant? Is the genome-wide GLR
  above or below 1? Together these tell you whether the circadian transcriptomes are
  related and in which direction.

# Function Quick Reference

| Step | Function | Key arguments |
|---|---|---|
| Initialize | `CB_init_single(Data.list, P=24)` | `Data.list = list(data, time, gname)` |
| MCMC | `CB_MCMC_single_rj_slice(Data.list, Init.value, ...)` | `iteration`, `n.burn`, `thin`, `p_rhythmic`, `diagnostics` |
| Annotate | `match_symbols(mcmc, BF=3)` | `BF` threshold for binary call |
| P(rhythmic) | `rowMeans(mcmc$rho)` | — |
| Bayes Factor | `summarize_bay(mcmc$rho, BF, p_rhythmic=0.2)` | `p_rhythmic` must match MCMC prior |
| Estimates+HDI | `CB_getAllEst(mcmc, burn=100)` | `credMass=0.95` |
| Transition | `transition_classify(pA, pB, bfdr_alpha=0.25)` | `bfdr_alpha` |
| Phase | `phase_infer(phi1, phi2, gain_loss_status, shift=2, ...)` | `shift`, `compute_hdi` |
| Pathway | `pathSelect(mcmc.merge.list, pathway.list, ranking.method, ...)` | `"union"`, `"gain"`, `"loss"`, `"conserved"` |
| Concordance | `multi_conservation(..., select.pathway.list="global")` | `n_perm`, `n_boot`, `save_output` |

# Citation

Pham T, *et al.* BayesRC: a comparative Bayesian multilevel framework for
evaluating circadian synchrony across conditions. *Manuscript in preparation*, 2025.
